### 2020-06-04 report/sections/server/node/runtime.tex
Report: complete chapter  4 runtime

abstract class IReceive {
	type: 'Receive' = 'Receive';
	abstract prepareReceive(
		next: EfsmTransitionHandler,
		register: (from: Roles.Peers, messageHandler: MessageHandler)
			=> void,
	): void;
};
\end{lstlisting}
\captionof{lstlisting}{\dots}
\label{lst:nodeabstractclass}
\end{figure}

@@

\begin{lstlisting}[language=javascript,tabsize=2]
// Inside Session class constructor...
this.next = this.next.bind(this);
this.send = this.send.bind(this);
this.register = this.register.bind(this);
\end{lstlisting}

@@

	performSend(
		next: EfsmTransitionHandler,
		send: (role: Roles.Peers, label: string, payload: any[]) 
			=> void
	) {
		const [label, payload, successor] = this.handler;
		send(Roles.Peers.Client, label, payload); (*@\label{line:nodesendrole}@*)
		return next(successor);
	}
}
\end{lstlisting}
\captionof{lstlisting}{\dots}
\label{lst:nodesend}
\end{figure}

@@

\item
When a message is received from the channel,
we dynamically process it with 
the current registered handler (\cref{line:callhandler}).
By MPST-safety, the incoming message must correspond
to the structure expected by the current receive state,
so correctness is preserved.
We encapsulate this dynamic behaviour in an instance method
and bind it as an event listener (\cref{line:wsbind})
for the WebSocket connection
of each non-server endpoint.
\end{enumerate}

@@

	receive({ data }: WebSocketMessage) {
		const handler = this.handler;
		this.handler = undefined;
		handler?.(data); (*@\label{line:callhandler}@*)
	}
}
\end{lstlisting}
\captionof{lstlisting}{Approach to Dynamic WebSocket Message Event Listener}
\label{lst:noderuntimewsmsg}
\end{figure}

@@

	prepareReceive(
		next: EfsmTransitionHandler,
		register: (from: Roles.Peers, messageHandler: MessageHandler)
			=> void
	) {
		const messageHandler = (message: any) => {
			const decoded = JSON.parse(message) as Message.S51;
			switch (decoded.label) {
				case Labels.S51.ADD: {
					const successor = 
						this.handler[decoded.label](...decoded.payload);
					return next(successor);
				}
				case Labels.S51.QUIT: {
					// Same as above...		
				}
			}            
		}
		register(Roles.Peers.Client, messageHandler);
	}
}
\end{lstlisting}
\captionof{lstlisting}{\dots}
\label{lst:nodereceive}
\end{figure}

Ideally, a more succinct (and direct) representation would be
\begin{lstlisting}[language=javascript,tabsize=2]
(message: any) => {
	const decoded = JSON.parse(message) as Message.S51;
	const successor =  this.handler[decoded.label](...decoded.payload);
	return next(successor);
}
\end{lstlisting}

But this expresses a type dependency between \texttt{label}
and \texttt{payload}, which as discussed 
(\cref{subsubsection:dependenttypes}), cannot be implemented.
However message structures \textit{precisely} 
define a discriminated union 
(the label acts as the discriminant 
to distinguish between payload types),
so we handle this with a switch statement,
at the cost of having the same code in each case --
TypeScript does infer the correct specific type in each case statement,
so code duplication here does serve a functional purpose.

@@

\[ A \to S: \tmsg{M1(string)}. B \to S: \tmsg{M2(number)}. \texttt{end} \]

Given that \trole{S} is the server endpoint,
we describe a possible execution flow for the protocol
that breaks our implementation:

\begin{enumerate}
\item 
\trole{S} transitions to its initial state, ``receive from \trole{A}''.
The receive handler for \tmsg{M1} is registered.

@@

\item
The \texttt{onmessage} handler for \tmsg{M2} is executed.
The registered handler expects \tmsg{M1(string)},
but it is callled with \tmsg{M2(number)}, which raises
a runtime type error.
\end{enumerate}

This exposes two problems: 
\textbf{(1)} the 
\end{example}

\begin{itemize}
\item motivate the edge case that message can arrive in the websocket (in succession) before the handler is registered
\item motivate the edge case that message can arrive `out of protocol order' 
\item explain the double queue system and how that provides consistency -- emphasising that this works because of the single-threaded typescript runtime
\end{itemize}

\begin{figure}[!ht]
\centering
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive2}
\caption{Message processed before transitioning to receive state}
\label{subfig:nodereceivemsgfirst}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive1}
\caption{Message processed after transitioning to receive state}
\label{subfig:nodereceivehandlefirst}
\end{subfigure}
\captionof{figure}{Possible Orderings for Handling Message Event and Preparing
Receive State}
\label{fig:nodereceivecompare}
\end{figure}

\subsection{Handling Termination}
WebSocket connections should be closed when the session terminates,
and both the browser endpoint and the server endpoint are capable
of closing connection. As we generate code for both, we define
a convention that the browser endpoint will initiate the WebSocket
close event.
As a result, we do nothing for this state at \cref{lst:noderuntime}.

>>>>

abstract class IReceive {
	type: 'Receive' = 'Receive';
	abstract prepareReceive(
		next: EfsmTransitionHandler,
		register: (from: Roles.Peers, messageHandler: MessageHandler)
			=> void,
	): void;
};
\end{lstlisting}
\captionof{lstlisting}{Class Definitions for \texttt{Implementation} API}
\label{lst:nodeabstractclass}
\end{figure}

@@

\begin{lstlisting}[language=javascript,tabsize=2]
// Inside the Session class constructor...
this.next = this.next.bind(this);
this.send = this.send.bind(this);
this.register = this.register.bind(this);
\end{lstlisting}

@@

	performSend(
		next: EfsmTransitionHandler,
		send: (role: Roles.Peers, label: string, payload: any[]) 
			=> void
	) {
		const [label, payload, successor] = this.handler;
		send(Roles.Peers.Client, label, payload); (*@\label{line:nodesendrole}@*)
		return next(successor);
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Code for \texttt{Implementation} API for
Send State}
\label{lst:nodesend}
\end{figure}

@@

\item
When a message is received from the channel,
we dynamically process it with 
the current registered handler (\cref{line:callhandler}).
We encapsulate this dynamic behaviour in an instance method
and bind it as an event listener (\cref{line:wsbind})
for the WebSocket connection
of each non-server endpoint.
\end{enumerate}

@@

	receive({ data }: WebSocketMessage) {
		const handler = this.handler; (*@\label{line:callhandlerstart}@*)
		this.handler = undefined; (*@\label{line:callhandler2}@*)
		handler?.(data); (*@\label{line:callhandler}@*)
	}
}
\end{lstlisting}
\captionof{lstlisting}{Approach to Dynamic WebSocket Message Event Listener}
\label{lst:noderuntimewsmsg}
\end{figure}

@@

	prepareReceive(
		next: EfsmTransitionHandler,
		register: (from: Roles.Peers, messageHandler: MessageHandler)
			=> void
	) {
		const messageHandler = (message: any) => {
			const decoded = JSON.parse(message) as Message.S51;
			switch (decoded.label) {
				case Labels.S51.ADD: {
					const successor = 
						this.handler[decoded.label](...decoded.payload);
					return next(successor);
				}
				case Labels.S51.QUIT: {
					// Same as above...		
				}
			}            
		}
		register(Roles.Peers.Client, messageHandler);
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Code for \texttt{Implementation} API for
Receive State}
\label{lst:nodereceive}
\end{figure}

Ideally, a more succinct (and direct) representation would be
\begin{lstlisting}[language=javascript]
(message: any) => {
	const decoded = JSON.parse(message) as Message.S51;
	const successor = 
		this.handler[decoded.label](...decoded.payload);
	return next(successor);
}
\end{lstlisting}

But this expresses a type dependency between \texttt{label}
and \texttt{payload} which, as discussed 
(\cref{subsubsection:dependenttypes}), cannot be implemented.
However message structures \textit{precisely} 
define a discriminated union 
(the label acts as the discriminant 
to distinguish between payload types),
so we handle this with a switch statement,
at the cost of having the same code in each case --
TypeScript does infer the correct specific type in each case statement,
so code duplication here does serve a functional purpose.

@@

\[
A \to S: \text{\lstonelinejs{M1(string)}}. 
B \to S: \text{\lstonelinejs{M2(number)}}. \texttt{end} 
\]

Suppose \trole{S} is the server endpoint.
We describe a possible execution flow for the protocol
that breaks our implementation:

\begin{enumerate}
\item 
\trole{S} transitions to its initial state, 
``receive \lstonelinejs{M1(string)} from \trole{A}''.
The receive handler for \tmsg{M1} is registered.

@@

\item
The \texttt{onmessage} handler for \tmsg{M2} is executed.
The registered handler expects \lstonelinejs{M1(string)},
but it is called with \lstonelinejs{M2(number)}, which raises
a runtime type error.
\end{enumerate}

This exposes a problem: 
the order of message \textit{arrivals} may not
correspond with the order of \textit{receiving} messages as
specified in the protocol,
\textit{so} the message may arrive before its
corresponding handler is registered.
\end{example}

We observe that message arrivals do not have to be causally related.
However, if we consider a similar \textit{binary} example
\[
A \to S: \text{\lstonelinejs{M1(string)}}. 
A \to S: \text{\lstonelinejs{M2(number)}}. \texttt{end}
\]
then \tmsg{M1} \textit{must} arrive before \textit{M2},
since this is sent through the WebSocket connection between 
\trole{A} and \trole{S}, and FIFO guarantees are respected 
for each individual WebSocket connection. 
We visualise the possible orders of message receive events
in \cref{fig:nodereceivecompare}.

\begin{figure}[!h]
\centering
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive2}
\caption{Message processed before transitioning to receive state}
\label{subfig:nodereceivemsgfirst}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive1}
\caption{Message processed after transitioning to receive state}
\label{subfig:nodereceivehandlefirst}
\end{subfigure}
\captionof{figure}{Possible Orderings for Handling Message Event and Preparing
Receive State}
\label{fig:nodereceivecompare}
\end{figure}

We observe that defining the handler using an optional type
is insufficient. We need a similar mechanism for handling messages
waiting for handlers, and we cannot assume an ordering on the arrival
of messages that are not causally related.

We proceed to generalise our approach from one optional-type
handler to two mappings:
\textbf{(1)} a mapping from endpoint to message queues\footnote{
TypeScript arrays have built-in $O(1)$ time complexity
\texttt{shift()} and \texttt{push()} operations, which
can be used as a queue.}, and
\textbf{(2)} a mapping from endpoint to handler queues.
To simply put, if an incoming message is waiting for its handler,
it gets enqueued in the message queue labelled by the sender of the message;
when the handler is created, it pops the message off the queue 
and directly processes it; the same logic applies for a handler waiting
for its message.
We could have used a mapping from endpoint to optional type,
but queue operations elegantly hides the mechanics of
\cref{line:callhandlerstart,line:callhandler2,line:callhandler}
in \cref{lst:noderuntimewsmsg}.

We outline the changes made to the \texttt{Session} class,
as shown in \cref{lst:nodesession}:

\begin{enumerate}
\item We construct types for these two mappings 
(\cref{line:mapped1,,line:mapped2}), 
using a generic mapped typed defined in
\cref{lst:noderolesmappedtypes}.

\item Empty queues are initialised for both mappings
in the constructor (\cref{line:queue1,,line:queue2}).

\item Each endpoint has a different \texttt{onmessage}
event listener, which will interact with the message queue
and handler queue corresponding to that endpoint. 
We achieve this by changing the \texttt{receive()} method
to be parameterised on the role instead 
(\cref{line:receivefunc}),
so it \textit{generates} an event listener 
(\cref{line:receivefuncret})
tailored for receiving messages
from that particular role.

\item
The \texttt{register()} method now also takes the role
as a parameter (\cref{line:newregister}) in order to check the 
corresponding pair of queues.
\end{enumerate}

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
type RoleToMessageQueue = Roles.PeersToMapped<any[]>; (*@\label{line:mapped1}@*)
type RoleToHandlerQueue = Roles.PeersToMapped<MessageHandler[]>; (*@\label{line:mapped2}@*)
type RoleToSocket = Roles.PeersToMapped<WebSocket>;

class Session {
	...
	private roleToSocket: RoleToSocket;
	private messageQueue: RoleToMessageQueue;
	private handlerQueue: RoleToHandlerQueue;

	constructor(...) {
		...
		Object.values(Roles.Peers).forEach(role => {
			const socket = this.roleToSocket[role];
			socket.onmessage = this.receive(role).bind(this);
		});

		this.messageQueue = { [Roles.Peers.Client]: [], }; (*@\label{line:queue1}@*)
		this.handlerQueue = { [Roles.Peers.Client]: [], }; (*@\label{line:queue2}@*)

		this.next(initialState);	
	}

	receive(from: Roles) { (*@\label{line:receivefunc}@*)
		return ({ data }) => { (*@\label{line:receivefuncret}@*)
			// Array.shift() can return undefined if empty
			const handler = this.handlerQueue[from].shift();
			if (handler !== undefined) {
				handler(data);
			} else {
				this.messageQueue[from].push(data);			
			}
		}
	}

	register(from: Roles.Peers, handler: MessageHandler) { (*@\label{line:newregister}@*)
		const message = this.messageQueue[from].shift();
		if (message !== undefined) {
			handler(message);		
		} else {
			this.handlerQueue[from].push(data);		
		}
	}
}
\end{lstlisting}
\captionof{lstlisting}{Modified \texttt{Session} class to correctly
handle message receive events}
\label{lst:nodesession}
\end{figure}

\begin{figure}[!h]
\begin{lstlisting}[language=javascript]
// Inside the Roles namespace...
export type PeersToMapped<Value> = { [Role in Peers]: Value };
\end{lstlisting}
\captionof{lstlisting}{Generic Mapped Type for Roles}
\label{lst:noderolesmappedtypes}
\end{figure}

Each endpoint must have consistency between
its message queue and handler queue.
We get the consistency from the simple fact that Node.js 
is a single-threaded runtime and execution is never pre-empted,
so there is no need to worry about atomic queue operations or
locking data structures.

We walk through how this design addresses the problems
in the previous example.

\begin{example}[Revisiting ``out-of-order'' message receives]
Consider the same multiparty session specified by the global type

\[
A \to S: \text{\lstonelinejs{M1(string)}}. 
B \to S: \text{\lstonelinejs{M2(number)}}. \texttt{end} 
\]

We show that our modified implementation addresses the problem:

\begin{enumerate}
\item 
\trole{S} transitions to its initial state,
``receive \lstonelinejs{M1(string)} from \trole{A}''.
\textit{The message queue for \trole{A} is empty,}
so the receive handler for \tmsg{M1} is registered 
\textit{under the handler queue for \trole{A}.}

\item
\tmsg{M2} arrives at \trole{S}, so the \texttt{onmessage}
handler is queued.

\item
The \texttt{onmessage} handler for \tmsg{M2} is executed.
\textit{The handler queue for \trole{B} is empty, 
so \tmsg{M2} is added to the message queue for \trole{B}}.

\item
The \texttt{onmessage} handler for \tmsg{M1} is executed.
\textit{The handler queue for \trole{A} is non-empty,
so the handler is popped off the front of the queue 
and processes \tmsg{M1}.}

\item
\trole{S} transitions to the successor state, 
``receive \lstonelinejs{M2(number)} from \trole{B}''.
\textit{The message for \trole{B} is non-empty,
so \tmsg{M2} is popped off the queue and 
processed by the handler.}
\end{enumerate}

This execution is free of communication mismatches.
\end{example}

\subsection{Handling Termination}
WebSocket connections should be closed when the session terminates,
and both the browser endpoint and the server endpoint are capable
of closing connection. As we generate code for both, we define
a convention that the browser endpoint will close 
the WebSocket connection,
so we do nothing for this state at \cref{lst:noderuntime}.
